---
title: "Lab 04 - Permutational Multivariate Tests"
author: "Nonparametric Statistics A.Y. 2024-2025"
date: 2024-10-24
output:
  html_document: 
    df_print: paged
    toc: true
  pdf_document: default
  word_document: default
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(rgl)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = T)
knitr::knit_hooks$set(webgl = hook_webgl)  # per il plot 3D
```

```{css, echo=FALSE}
.extracode {
background-color: lightblue;
}
```



```{r}
B = 5e3
seed = 20
```


## Permutational Multivariate Tests

In this part of the lab, we will directly implement some permutational multivariate tests, applied to specific problems
Let's start with our first one...

### Two multivariate populations test

We are given data of the week 12/09/2016 -> 18/09/2016 about the number of vehicles that, **each half-hour**, enter Milan Low emission zone (the so called "_area C_"). You want to know if the mean entrance pattern in the weekends is significantly different from the one in the weekdays.
Framed in a more rigorous way, being 

$$
\mathbf{y}_{i,w} \overset{iid}{\sim} \mathbf{Y}_{w} \in \mathbb{R}^{48}; \, i \in \{1,...,5\}
$$ 
where $w$ denotes the distribution of weekdays, and $i$ is the $i$th day of the week. We also have:
$$
\mathbf{y}_{i,\tilde{w}} \overset{iid}{\sim} \mathbf{Y}_{\tilde{w}} \in \mathbb{R}^{48} \; i \in \{6,7\}
$$
where $\tilde{w}$ is to denote weekend day.
We want to test the equality of the two distributions, namely, we want to devise a test of the type:
$$
H_0: \mathbf{Y}_{w} \overset{d}{=} \mathbf{Y}_{\tilde{w}}\;vs\;H_1:\mathbf{Y}_{w} \overset{d}{\neq} \mathbf{Y}_{\tilde{w}}
$$

Let's start by reading and rearranging the data.

```{r}
# set header = TRUE read the first row in the csv as colnames.
# for every day, we will have observations corresponding to every half hour
d1 = read.csv('./data/areac_data/accessi-orari-areac-2016-09-12-00_00_00.csv', header=T)
d2 = read.csv('./data/areac_data/accessi-orari-areac-2016-09-13-00_00_00.csv', header=T)
d3 = read.csv('./data/areac_data/accessi-orari-areac-2016-09-14-00_00_00.csv', header=T)
d4 = read.csv('./data/areac_data/accessi-orari-areac-2016-09-15-00_00_00.csv', header=T)
d5 = read.csv('./data/areac_data/accessi-orari-areac-2016-09-16-00_00_00.csv', header=T)
d6 = read.csv('./data/areac_data/accessi-orari-areac-2016-09-17-00_00_00.csv', header=T)
d7 = read.csv('./data/areac_data/accessi-orari-areac-2016-09-18-00_00_00.csv', header=T)

# we bind the rows corresponding to every day.
week = rbind(d1[,2], d2[,2], d3[,2], d4[,2], d5[,2],
             d6[,2], d7[,2])
# plot
matplot(seq(0,47)/2, # every half-hour of the day
        t(week), # we want to plat the columns
        type='l',
        col=c(1,1,1,1,1,2,2), # colour weekdays and weekend-days differentlyy
        lty=1)

```

As you remember, we can actually choose whatever test statistic we may like: if the permutation scheme used during the test is likelihood-invariant, we will get in any case an exact test.
There are nevertheless better choices than others (in the sense that we have a higher power, as we saw in the previous lab).
Let's try to use the squared euclidean distance between the two sample mean vectors (admittedly a quite standard choice.)

```{r}
t1 = week[1:5,]
t2 = week[6:7,]

t1.mean = colMeans(t1)
t2.mean = colMeans(t2)

matplot(seq(0,47)/2,
        t(rbind(t1.mean,t2.mean)), 
        type='l', col=c(1,2), lty=1,
        main="Sample multivariate means")
```

Let's compute the test statistic $T_0 =||\bar{\mathbf{x}}_1 -\bar{\mathbf{x}}_2||_2^2 =(\bar{\mathbf{x}}_1-\bar{\mathbf{x}}_2)^\intercal(\bar{\mathbf{x}}_1-\bar{\mathbf{x}}_2) $ where $||.||_2$ denotes the euclidean norm in $\mathbb{R}^{48}$.

```{r}
n1 = dim(t1)[1]
n2 = dim(t2)[1]
n  = n1 + n2

T20 = as.numeric(t(t1.mean-t2.mean) %*% (t1.mean-t2.mean))  # matrix (vector) product
T20
```

To perform our test, we need to compare the test statistic to its (permutational) distribution under $H_0$.

```{r}
# Estimating the permutational distribution under H0

T2 = numeric(B)
set.seed(seed)
for(perm in 1:B){
  # Random permutation of indexes
  # When we apply permutations in a multivariate case, we keep the units together
  # i.e., we only permute the rows of the data matrix
  t_pooled = rbind(t1,t2)
  permutation = sample(n)
  t_perm = t_pooled[permutation,]
  t1_perm = t_perm[1:n1,]
  t2_perm = t_perm[(n1+1):n,]
  
  # Evaluation of the test statistic on permuted data
  t1.mean_perm = colMeans(t1_perm)
  t2.mean_perm = colMeans(t2_perm)
  T2[perm]  = t(t1.mean_perm-t2.mean_perm) %*% (t1.mean_perm-t2.mean_perm) 
}
```

Let's now see the shape of the permutational distribution, compared with the computed test statistic (the green vertical line...)

```{r}
hist(T2,xlim=range(c(T2,T20)))
abline(v=T20,col=3,lwd=4)

plot(ecdf(T2))
abline(v=T20,col=3,lwd=4)
```

The P-value will be amazingly low... but let's try to calculate it nevertheless.
Recall the p-value is given by:
$$
p = B^{-1}\sum_{b=1}^B\mathbb{I}_{\{T_b \geq T_0 \}}
$$

```{r}
p_val = sum(T2>=T20)/B
p_val
```

With this P-value, I can say, with a level of confidence higher than $95 \%$ that weekdays and weekends are significantly different

let's see another case now...


### Center of symmetry permutation tests

#### Multivariate case


We're still in Milan, but now we want to check if the humidity of ($4$) summer months is significantly different from the "non-comfort" threshold level of $(65\%)$. _Id est_,  are interested in the **location** of a (multivariate) distribution.

In other words, being ($i$ represents the sample's $i$th datum )
$$
\mathbf{y}_{i} \overset{iid}{\sim} \mathbf{Y} \in \mathbb{R}^{4}
$$
, I want to test

$$
H_0: \mathbb{E}[\mathbf{Y}] = \mathbf{\mu}_{0}\;vs\;H_1:\mathbb{E}[\mathbf{Y}] \neq \mathbf{\mu}_0
$$
Let's read and plot the data also here.
We will have 4 observations ($p=4$) for 7 different year ($n=7$).

```{r}
hum = read.csv2('./data/humidity_data/307_Umidita_relativa_2008_2014.csv', header=T)
hum = hum$Media  # we use only the mean humidity
hum = matrix(hum,  # cast as matrix
             ncol=12, byrow=T)[,6:9] # so we have n = 7 and p =4

boxplot(hum)
matplot(t(hum),
        type='l',
        lty=1,
        main="Humidity at different months, at different years (colours)",
        xlab="humidity",
        ylab="obs.month")

```

What is a reasonable permutation scheme to implement, and thus what is a reasonable test to perform? It is actually harder with respect to two-sample tests... but if we **assume the distribution to be symmetric**, reflections are permutationally invariant, and thus I can easily test this! Let's define the center of symmetry

```{r}
mu0      = c(65, 65, 65, 65)
```

Let's compute the test statistic (here the squared distance between the sample mean and the hypothesised centre, but of course other choices are possible...)


```{r}
x.mean   = colMeans(hum)
n = dim(hum)[1]
p = dim(hum)[2]

T20 = as.numeric(t(x.mean-mu0) %*% (x.mean-mu0) )
```


And now, let's compute the permutational distribution!

```{r}
T2 = numeric(B) 
set.seed(20)

for(perm in 1:B){
  # In this case we use changes of signs in place of permutations
  
  # Permuted dataset (reflection-based)
  signs.perm = rbinom(n, 1, 0.5)*2 - 1  
  hum_perm = mu0 + (hum - mu0) * matrix(signs.perm, nrow=n,ncol=p,byrow=FALSE)  # hadamard product
  x.mean_perm = colMeans(hum_perm)
  T2[perm]  = t(x.mean_perm-mu0)  %*% (x.mean_perm-mu0) #n.b. also works without the t, but better to play it safe!
}
```

let's plot the permutational distribution of the test statistic

```{r}
hist(T2,xlim=range(c(T2,T20)))
abline(v=T20,col=3,lwd=4)

plot(ecdf(T2))
abline(v=T20,col=3,lwd=4)
```

and the p-value

```{r}
p_val <- sum(T2>=T20)/B
p_val

```

Also here, I can argue that the humidity during the summer months is, with a 95% level of confidence ( _i.e._ 5% of the times I will reject $H_0$ when it is actually true), significantly different from $65%$

#### Univariate case

Of course this generalises the univariate case. We might take the observed humidity at only one month (univariate), and different years are (independent) observations.

In terms of code, this is looking at one column of the matrix.



```{r}
x.mean  <- mean(hum[,1])
mu.0 <- 65
n = dim(hum)[1]
p = 1

T20 = (x.mean - mu.0)^2

T2 = numeric(B) 
set.seed(20)

for(perm in 1:B){
  # In this case we use changes of signs in place of permutations
  
  # Permuted dataset (reflection-based)
  signs.perm = rbinom(n, 1, 0.5)*2 - 1  
  hum_perm = as.vector(mu.0 + (hum[,1] - mu.0) * matrix(signs.perm, nrow=n,ncol=1,byrow=FALSE))  # hadamard product
  x.mean_perm = mean(hum_perm)
  T2[perm]  = (x.mean_perm-mu.0)**2
}

```

And we look at the (MC-estimated) permutational distribution of the test statistic under $H_0$

```{r}
hist(T2,xlim=range(c(T2,T20)))
abline(v=T20,col=3,lwd=4)

plot(ecdf(T2))
abline(v=T20,col=3,lwd=4)
```

and the p-value

```{r}
p_val <- sum(T2>=T20)/B
p_val

```


### Two sample paired multivariate permutation test


In this case, we want to compare the temperature, humidity and wind-speed in 50 different days in Milan and Barcelona.
We deem $X_M$ the r.v. for the quantity of interest for the population of Milan with unknown distribution $F_M$ and $X_B$ with $F_B$ those of Barcelona.
We have data:
$$
\mathbf{x}_i \stackrel{iid}{\sim} F_M; i \in \{1, ..., N\}
$$
_et_
$$
\mathbf{y}_i \stackrel{iid}{\sim} F_B; i \in \{1, ..., N\}
$$
And, very importantly, the observations are paired:
$$
(\mathbf{x}_i - \mathbf{y}_i) \stackrel{iid}{\sim} \tilde{F}
$$
for some unknown bivariate distribution $\tilde{F}$

#### Option 1


We test:
$$
H_0 : \mathbb{E}[X_M-X_B] = \mathbf{0} \, versus \, H_1: \mathbb{E}[X_M-X_B] \neq \mathbf{0}
$$

Let's read the data
```{r}
t1 <- read.table('./data/meteo_data/barcellona.txt', header=T)
t2 <- read.table('./data/meteo_data/milano.txt', header=T)

```

Let's try to explore the data... we can work with the paired differences and plot them.

```{r}
library(rgl)
open3d()
plot3d(t1-t2, size=3, col='orange', aspect = F)
points3d(0,0,0, size=6)

p  <- dim(t1)[2]
n1 <- dim(t1)[1]
n2 <- dim(t2)[1]
n <- n1+n2

```

In terms of permutation schemes (and testing strategies...) to follow, the best choice is to compute the differences between the two groups, assume their distribution to be symmetric, and then perform a centre of symmetry test.

What is the best test statistic for the test? let's see...

```{r}
t1.mean <- colMeans(t1)
t2.mean <- colMeans(t2)
t1.cov  <-  cov(t1)
t2.cov  <-  cov(t2)
Sp      <- ((n1-1)*t1.cov + (n2-1)*t2.cov)/(n1+n2-2)  # pooled cov matrix
Spinv   <- solve(Sp)

delta.0 <- c(0,0,0)

diff <- t1-t2
diff.mean <- colMeans(diff)
diff.cov <- cov(diff)
diff.invcov <- solve(diff.cov)
```

Let's start with the squared euclidean distance between the difference in means and the hypothesised value

```{r}
T20 <- as.numeric(t(diff.mean-delta.0)  %*% (diff.mean-delta.0))
```

And then, let's perform the test

```{r}
T2 <- numeric(B)
set.seed(seed)
for(perm in 1:B)
  {
  # Random permutation
  # obs: exchanging data within couples means changing the sign of the difference
  signs.perm <- rbinom(n1, 1, 0.5)*2 - 1
  
  diff_perm <- diff * matrix(signs.perm,nrow=n1,ncol=p,byrow=FALSE)
  diff.mean_perm <- colMeans(diff_perm)
  diff.cov_perm <- cov(diff_perm)
  diff.invcov_perm <- solve(diff.cov_perm)
  
  T2[perm] <- as.numeric(t(diff.mean_perm-delta.0) %*% (diff.mean_perm-delta.0))
  }
```
Distribution and pvalue

```{r}
# plotting the permutational distribution under H0
hist(T2,xlim=range(c(T2,T20)),breaks=100)
abline(v=T20,col=3,lwd=4)

plot(ecdf(T2))
abline(v=T20,col=3,lwd=4)


# p-value
p_val <- sum(T2>=T20)/B
p_val

```



Now, let's use the Mahalanobis distance, but "forgetting" about the covariance between the values

```{r}
T20 <- as.numeric( t(diff.mean-delta.0) %*% solve(diag(diag(diff.cov))) %*% (diff.mean-delta.0))
# Estimating the permutational distribution under H0
T2 <- numeric(B)
set.seed(seed)
for(perm in 1:B)
  {
  # Random permutation
  # obs: exchanging data within couples means changing the sign of the difference
  signs.perm <- rbinom(n1, 1, 0.5)*2 - 1
  
  diff_perm <- diff * matrix(signs.perm,nrow=n1,ncol=p,byrow=FALSE)
  diff.mean_perm <- colMeans(diff_perm)
  diff.cov_perm <- cov(diff_perm)
  diff.invcov_perm <- solve(diff.cov_perm)
  

  T2[perm] <- as.numeric((diff.mean_perm-delta.0) %*% solve(diag(diag(diff.cov_perm))) %*% (diff.mean_perm-delta.0))
  
}

# plotting the permutational distribution under H0
hist(T2,xlim=range(c(T2,T20)),breaks=100)
abline(v=T20,col=3,lwd=4)

plot(ecdf(T2))
abline(v=T20,col=3,lwd=4)


# p-value
p_val <- sum(T2>=T20)/B
p_val

```

and lastly, let's use the proper Mahalanobis distance

```{r}
T20 <- as.numeric((diff.mean-delta.0) %*% diff.invcov %*% (diff.mean-delta.0))



# Estimating the permutational distribution under H0

set.seed(seed)
T2 <- numeric(B)

for(perm in 1:B)
  {
  # Random permutation
  # obs: exchanging data within couples means changing the sign of the difference
  signs.perm <- rbinom(n1, 1, 0.5)*2 - 1
  
  diff_perm <- diff * matrix(signs.perm,nrow=n1,ncol=p,byrow=FALSE)
  diff.mean_perm <- colMeans(diff_perm)
  diff.cov_perm <- cov(diff_perm)
  diff.invcov_perm <- solve(diff.cov_perm)
  
  #T2[perm] <- as.numeric(n1 * (diff.mean_perm-delta.0) %*% (diff.mean_perm-delta.0))
  #T2[perm] <- as.numeric(n1 * (diff.mean_perm-delta.0) %*% solve(diag(diag(diff.cov_perm))) %*% (diff.mean_perm-delta.0))
  T2[perm] <- as.numeric(n1 * (diff.mean_perm-delta.0) %*% diff.invcov_perm %*% (diff.mean_perm-delta.0))
  }

# plotting the permutational distribution under H0
hist(T2,xlim=range(c(T2,T20)),breaks=100)
abline(v=T20,col=3,lwd=4)

plot(ecdf(T2))
abline(v=T20,col=3,lwd=4)


# p-value
p_val <- sum(T2>=T20)/B
p_val

```


#### Option 2

If you read the _dispensa_ created by Professor Alessia Pini^[Unpublished, _cfr._ the material on Webeep.], you would find an alternative for this test. 

Instead of performing inference on the distribution of 
$$
X_M - X_B
$$
we treat them "separately": $(X_M, X_B) \sim F$ for some **bivariate** distribution $F$. And of course we require likelihood-invariant transformations under $H_0$
that respect the paired nature of the test.
We test:
$$
H_0: X_M \stackrel{d}{=} X_B \; vs. \;H_1: X_M \stackrel{d}{\neq} X_B
$$
And we have $2^N$ different permutations, since the **exchangeability is only within the pairs**.
We choose as test statistic the norm of the differences of the means of both samples.


```{r}
T20 <- norm(as.matrix(colMeans(t1) - apply(t2, MARGIN=2, FUN=mean)))
```


```{r}
T2 <- numeric(B)
p <- dim(t1)[2] # here naturally I can use t1 or t2.
n <- dim(t2)[1]
t.full <- rbind(t1, t2)
set.seed(seed)
for(perm in 1:B)
  {
  # Random permutation
  # N.B. exchangeability is only within pairs
  perm.indices.t1 <- seq(1, n) + n * rbinom(n,1, 0.5)
  t1.perm <- t.full[perm.indices.t1, ]
  t2.perm <- t.full[-perm.indices.t1,]
  
  T2[perm] <- norm(as.matrix(((colMeans(t1.perm)) - colMeans(t2.perm))))
}

hist(T2,xlim=range(c(T2,T20)),breaks=100)
abline(v=T20,col=3,lwd=4)

plot(ecdf(T2))
abline(v=T20,col=3,lwd=4)


# p-value
p_val <- sum(T2>=T20)/B
p_val
```

## Permutational regression: categorical regressors

### Permutational ANOVA

Until now, we have seen how to perform basic statistical tests in a
nonparametric and, mainly and more importantly, in a permutational
framework. By going further on this line of thinking, one can perform
analysis of variance in a fully permutational (and thus exact...)
setting. The example we will see today is about a fairly old dataset,
appeared on Biometrika in 1948 (!!). It is about an experiment over a
number of chickens, fed with different feed types, which were weighted
by the experimenter. You are interested in determining if the feed type
has an impact on the chicken weight or not.

Let's import and summarise the dataset

```{r}
chickwts
attach(chickwts)
summary(chickwts)
```

and, let's try to plot the results, to get a bit of info about what's
going on...

```{r}
g <- nlevels(feed)
n <- dim(chickwts)[1]

plot(feed, weight, xlab='treat',col=rainbow(g),main='Original Data')
```

The null hypothesis that we want to test is that, being
$\tau_i, i\in \{1,\ldots6\}$ the generic effect of a given level of the
treatment.
$$
H_0: \tau_i=0\;\forall i\;vs\;H_1:\exists\,\tau_i\neq0 
$$

Of course we can solve this in a fully parametric (and admittedly not
robust...) way

```{r}
fit <- aov(weight ~ feed)
summary(fit)

```

The strategy that I use to perform permutational ANOVA is to
"permutationalise" the F statistic, by computing its permutational
distribution under $H_0$. So...

```{r}
T0 <- summary(fit)[[1]][1,4]  # extract the test statistic
T0

```

To compute the distribution, one simply "scales up" the one used for the
2-sample t-test: I assign at random the treatments (that, under $H_0$,
should all be equal, and equal to 0)

```{r}
T_stat <- numeric(B) 
n <- dim(chickwts)[1]

for(perm in 1:B){
  # Permutation:
  permutation <- sample(1:n)
  weight_perm <- weight[permutation]
  fit_perm <- aov(weight_perm ~ feed)
  
  # Test statistic:
  T_stat[perm] <- summary(fit_perm)[[1]][1,4]
}

```

Let's see the distribution, and then the p-value of the permutational
f-test

```{r}
hist(T_stat,xlim=range(c(T_stat,T0)),breaks=30)
abline(v=T0,col=3,lwd=2)

plot(ecdf(T_stat),xlim=c(-1,20))
abline(v=T0,col=3,lwd=4)

# p-value
p_val <- sum(T_stat>=T0)/B
p_val
```


**It can be verified in this case the residual permutation and the data
permutation are the same**...

```{r}
T0.fit <- summary(fit)[[1]][1,4]

# residuals under H0 (i.e all tau = 0)
# weight = mu
glob.intercept <- mean(weight)
residuals.H0 <- weight - glob.intercept # subtract global mean
residuals.H0.aov <- (aov(weight ~ 1))$residuals  # alternative: code model under H0
all.equal(residuals.H0, as.vector(residuals.H0.aov))

# Note that in this case, permuting the residuals under H0 
# and permuting the data is exactly the same:
permutation <- sample(n)
weight.perm.res <- glob.intercept + residuals.H0[permutation]
weight.perm        <- weight[permutation]

all.equal(weight.perm.res, weight.perm)
```

Indeed, with such _perspicacité_, we realise:

#### POV: ANOVA is a special case of a linear model.

Recall we are building a model (under $H_1$)
$$
weight_i = \beta_0 + \tau_{j_i} + \epsilon_i\,\,\,, j \in \{1,\ldots6\}
$$
whereas under $H_0$ all $\tau_i=0$, so we have
$$
weight_i = \beta_0 + \epsilon_i
$$
  Under the null hypothesis the model **permuting the residuals** should yield
likelihood-invariant datasets. However, instead of using the $aov$
function, we could use $lm$! And moreover we now have access to the
adjusted R squared statistic (among others!).

```{r}
typeof(feed)  # even if we have names, since it is a factor
```

We do the test with the adjusted R squared statistic

```{r}
T0.chicks <- summary.lm(lm(weight ~ feed))$adj.r.squared
T0.chicks


lm.H0.chicks <- lm(weight ~ 1)
lm.H0.chicks
residuals.H0.chicks <- lm.H0.chicks$residuals
n = length(weight)

T_chicks <- numeric(B)
set.seed(seed)
for(perm in 1:B){
  permutation <- sample(n)
  residuals.H0.perm <- residuals.H0.chicks[permutation]
  weight.perm.H0<- lm.H0.chicks$fitted + residuals.H0.perm
  T_chicks[perm] <- summary.lm(lm(weight.perm.H0~ feed))$adj.r.squared
}

sum(T0.chicks <= T_chicks)/B


```

Or the F statistic (same as before)
```{r}
# model under H1
T0
T0 <- summary.lm(lm(weight ~ feed ))$f[1]
T0
T_chicks <- numeric(B)
set.seed(seed)
for(perm in 1:B){
  permutation <- sample(n)
  weight.perm <- weight[permutation]
  T_chicks[perm] <- summary.lm(lm(weight.perm ~ feed ))$f[1]
  # T_fuel[perm] <- summary.lm(aov(weight.perm ~ feed ))$f[1] # same
  
}
sum(T_chicks >= T0)/B

```


#### Why you should be diffident of unknown packages

Let's see what happens instead when using `lmPerm`

```{r, eval=F}
library(lmPerm)
lmp=aovp(weight_perm ~ chickwts$feed,
         perm="Prob",
         Cp=0.1)#cp is supposed to stop iterations when standard error is at that level...
summary(lmp)
```

let's reduce the expected standard error...

```{r, eval=F}
lmp=aovp(weight_perm ~ chickwts$feed,perm="Prob",Cp=1e-6)
summary(lmp)
```

and let's run the instruction several times

```{r, eval=F}
lmp=aovp(weight_perm ~ chickwts$feed,perm="Prob",Cp=1e-6)
summary(lmp)

lmp=aovp(weight_perm ~ chickwts$feed,perm="Prob",Cp=1e-6)
summary(lmp)

lmp=aovp(weight_perm ~ chickwts$feed,perm="Prob",Cp=1e-6)
summary(lmp)

```

One may hypothesise such variance in the obtained p-value is a
consequence of the number of iterations. Anyways, **it doesn't really
make much sense. Don't use it.**


Up to now, everything is fairly easy. The situation gets a bit uglier
when more than one factor is involved in the model

### Permutational Two-Way ANOVA

That is, multiple regression with categorical regressors.


In this case we wan to test the efficiency of a car with different
fuels, which are classified by their **producer** or by their **octane**
number. I want to see what are the meaningful factors in determining the
number of kilometers performed with a liter of petrol, interaction
included...

This time the data is put in by hand...

```{r}
km          <- c(18.7, 16.8, 20.1, 22.4, 14.0, 15.2, 22.0, 23.3)
station     <- factor(c('Esso','Esso','Esso','Esso','Shell','Shell','Shell','Shell'))
fuel        <- factor(c('95','95','98','98','95','95','98','98'))
station_fuel<- factor(c('Esso95','Esso95','Esso98','Esso98','Shell95','Shell95','Shell98','Shell98'))

M             <- mean(km)
Mstation      <- tapply(km,      station, mean)
Mfuel         <- tapply(km,       fuel, mean)
Mstation_fuel <- tapply(km, station_fuel, mean)
```

Let's also plot the data

```{r}
plot(station_fuel, km, col=rainbow(5)[2:5], ylim=c(0,24))
```

and, of course, I can do everything in a fully parametric setting...

```{r}
# Parametric test:
summary(aov(km ~ station + fuel + station:fuel))
# Without interaction
summary.aov(aov(km ~ station + fuel))
# Without station
summary.aov(aov(km ~ fuel))

```

**To do everything in a permutational setting, I need to recognize that,
actually, the tests that I need to run are more than one!**

The two-way ANOVA full model is: $$
Km = \mu + \alpha_i + \beta_j + \gamma_{ij} + \epsilon
$$

So, I need to see if $\gamma=0$, or $\beta=0$ or $\alpha=0$ and, for
each of this tests, I have a different permutation scheme.

Let's start with the $$
H_0:\gamma=0 \; vs \; H_1:\gamma\neq0
$$

Which means that the model under $H_1$ is given by: $$
Km = \mu + \alpha_i + \beta_j + \gamma_{ij} +  \epsilon
$$ And under $H_0$: $$
Km = \mu + \alpha_i + \beta_j + \epsilon
$$ , and **permuting the residuals** under this model should yield
likelihood-invariant datasets. So, let's compute the test statistic

```{r}
summary.aov(aov(km ~ station + fuel + station:fuel)) 
T0_station_fuel <- summary.aov(aov(km ~ station + fuel + station:fuel))[[1]][3,4]  # extract the test statistic
T0_station_fuel

```

and compute the permutational distribution. Note that we have to
**estimate** the residuals which we will permute.

```{r}
aov.H0station_fuel <- aov(km ~ station + fuel)
aov.H0station_fuel
residuals.H0station_fuel <- aov.H0station_fuel$residuals # estimate residuals
n = 8


T_station_fuel <- numeric(B)
for(perm in 1:B){
  permutation <- sample(n)
  # permute the residuals
  residuals.H0station_fuel <- residuals.H0station_fuel[permutation]
  km.perm.H0station_fuel <- aov.H0station_fuel$fitted + residuals.H0station_fuel
  # re-fit full model to calculate statistic
  T_station_fuel[perm] <- summary.aov(aov(km.perm.H0station_fuel ~ station + fuel + station:fuel))[[1]][3,4]
}
```

And the p-value...

```{r}
sum(T_station_fuel >= T0_station_fuel)/B
```

Not significant, meaning that I can reduce the model and then perform my
test on my main effects.

To test $H_0:\beta=0$ vs $H_1:\beta\neq 0$, I am assuming under $H_0$
the following model $Km = \mu + \alpha_i + \epsilon$, while for
$H_0:\alpha=0$ vs $H_1:\alpha\neq0$  I am assuming
$Km = \mu + \beta_j + \epsilon$ under $H_0$. That is, in the next code
blocks I will perform **two separate tests\**.

Again, the idea is to permute the residuals under $H_0$, so let's
compute them.

```{r}
# Test for station
T0_station <- summary.aov(aov(km ~ station + fuel))[[1]][1,4]
# residuals under H0:
# km = mu + beta*fuel
aov.H0station <- aov(km ~ fuel)
residuals.H0station <- aov.H0station$residuals

# Test for fuel
T0_fuel <- summary.aov(aov(km ~ station + fuel))[[1]][2,4]
# residuals under H0:
# km = mu + alpha*station
aov.H0fuel <- aov(km ~ station)
residuals.H0fuel <- aov.H0fuel$residuals

```

And let's now compute the permutational distribution, and thus the
p-value

```{r}
# Test both factors in a single loop
B <- 1000
T_fuel <- T_station <- numeric(B)
for(perm in 1:B){
  permutation <- sample(n)
  
  # Test station
  km.perm.H0station <- aov.H0station$fitted + residuals.H0station[permutation]
  T_station[perm] <- summary.aov(aov(km.perm.H0station ~ station + fuel))[[1]][1,4]
  
  # Test fuel
  km.perm.H0fuel <- aov.H0fuel$fitted + residuals.H0fuel[permutation]
  T_fuel[perm] <- summary.aov(aov(km.perm.H0fuel ~ station + fuel))[[1]][2,4]
}
```

```{r}
sum(T_station >= T0_station)/B
```

```{r}
sum(T_fuel >= T0_fuel)/B
```

In histograms

```{r}
hist(T_station)
abline(v=T0_station)

hist(T_fuel)
abline(v=T0_fuel)
```

I can remove station, so let's go on on testing Fuel... which actually
is a one-way ANOVA... 
Hence, we have our last hypothesis test: $$
H_0:  \beta = 0 \implies Km = \mu + \epsilon
$$ *versus* $$
H_1:  \beta \neq 0 \implies Km = \mu + \beta_j + \epsilon
$$ **It can be shown in this case the residual permutation and the data
permutation are the same**...

```{r}
# TEST ON THE FACTOR FUEL
T0_fuel <- summary.aov(aov(km ~  fuel))[[1]][1,4]
# residuals under H0
# km = mu
residuals.H0fuel <- km - M  # subtract global mean

# Note that in this case, permuting the residuals under H0 
# and permuting the data is exactly the same:
permutation <- sample(n)
km.perm.H0fuel <- M + residuals.H0fuel[permutation]
km.perm        <- km[permutation]

all.equal(km.perm.H0fuel, km.perm)
```

```{r}
T_fuel <- numeric(B)
for(perm in 1:B){
  permutation <- sample(n)
  km.perm <- km[permutation]
  T_fuel[perm] <- summary.lm(aov(km.perm ~ fuel ))$f[1]
  
}
sum(T_fuel >= T0_fuel)/B

```

### Multivariate Analysis of Variance

The strategy behind the multivariate analysis of variance is admittedly
fairly similar. We use the Iris dataset

```{r}
data(iris)
attach(iris)
head(iris)
table(iris$Species)
```

let's arrange it a bit, and plot it.

```{r}
species.name <- factor(Species, labels=c('setosa','versicolor','virginica'))
iris4        <- iris[,1:4]
plot(iris4,col=species.name)
```

```{r}
i1 <- which(species.name=='setosa')
i2 <- which(species.name=='versicolor')
i3 <- which(species.name=='virginica')
n1 <- length(i1)
n2 <- length(i2)
n3 <- length(i3)
n  <- n1+n2+n3

g  <- length(levels(species.name))
p  <- 4
```

How to perform a MANOVA test? instead of using the F-test, we will
develop a test based on a "permutationalisation" of the Wilks[^1] test.
Let's compute the permutational test statistic

[^1]: I refer you to the textbook of Applied Statistics for info on this
    statistic

```{r}
fit <- manova(as.matrix(iris4) ~ species.name)
summary.manova(fit,test="Wilks") 
T0 <- -summary.manova(fit,test="Wilks")$stats[1,2]
T0
```

And let's now compute instead the permutational distribution of the test
statistic.

```{r}
set.seed(seed)
T_stat <- numeric(B)

for(perm in 1:B){
  # choose random permutation
  permutation <- sample(1:n)
  species.name.perm <- species.name[permutation]
  fit.perm <- manova(as.matrix(iris4) ~ species.name.perm)
  T_stat[perm] <- -summary.manova(fit.perm,test="Wilks")$stats[1,2]
}
```

Let's visualize again the distribution, and the p-value

```{r}
hist(T_stat,xlim=range(c(T_stat,T0)),breaks=30)
abline(v=T0,col=3,lwd=2)

plot(ecdf(T_stat),xlim=c(-2,1))
abline(v=T0,col=3,lwd=4)

# p-value
p_val <- sum(T_stat>=T0)/B
p_val
```


## Permutational regression: continuous regressors

### Simple regression

We have now seen many applications of permutation testing to various real-world methodological issues... The last one is regression.
In this case we will work with some simulated data: specifically data generated from a linear model with a **heavy-tailed error term** (we have seen in **Lab$03$** the effect of heavy tails of a distribution in the _t-test_).

Here is a recap of the algorithm you have studied in the lectures.


******
**Algorithm 1**:  Permutation test algorithm for linear models

******
1.  Set $H_0$ (smaller model) and $H_1$ (more complex model). _E.g._, 
$$
H_1: y_i = \beta_0 + \beta_1 x_{i1} + \epsilon_i
$$
$$
H_0: y_i = \beta_0 + \epsilon_i
$$
    where the residuals are of an unknown distibution, but we do assume they are i.i.d with $\mathbb{E}[\epsilon]=0$.
2. Fit the  full model ($H_1$), and extract the value of test statistic (_e.g._, the $F$ statistic)
3. Fit the reduced model ($H_0$ )to estimate the residuals, which are exchangeable under $H_0$. Extract as well the fitted values $\hat{y}_i$. 
4. Run Conditional Monte Carlo simulation to estimate the permutational distribution of the statistic conditional on the sample. For example, we may have in the $k$ iteration, the $k$ permutational sample, and for each $i$:
$$
y_i^{(k)} = \hat{y}_i + \epsilon_i^{(k)}
$$
Where $\hat{y}_i$ is the fitted value of the reduced model in step ($2$) and $\epsilon_i^{(k)}$ a sample without replacement of the estimated residuals. With such dependent variable, we fit the full model ( _i.e._ the model under $H_1$) and extract $T^{(k)}$, the value of the test statistic in the $k$th iteration )

5. Compare the value of the statistic in ($2$) with the permutational distribution estimated in ($4$) of $T^{(k)}, \; k \in \mathcal{K}$to yield the p-value.

******


```{r}
set.seed(seed)
# covariate values
x1 <- runif(n,0,10)

# generating model
b0 <- 2
b1 <- 3
Y <- b0 + b1*x1 + stabledist::rstable(n,1.2,0)  # heavy tailed term
```

```{r}
plot(x1,Y,pch=16)
```

```{r}
# model under H_1
result <- lm(Y ~ x1)
summary(result)
```

One may ask: why use a nonparametric test?

We test the residuals for normality
```{r}
shapiro.test(result$residuals)
```

**Exercise**: show through a DD plot there is no match between the distribution of th (estimated) residuals  and a normal one (more on this below).

We proceed by applying **Algorithm 1**.

$$
H_0: y_i = \beta_0 + \epsilon_i \,\,\, versus \,\,\,\ H_1: y_1 = \beta_0 + \beta_1 x_i + \epsilon_i
$$

```{r}
# observed test statistics
T0.beta.sq <- (summary.lm(result)$coefficients[2,1])^2
# in simple regression, same as R squared
# T0.beta <- (summary.lm(result)$r.squared)


T0.beta.std <- (summary.lm(result)$coefficients[2,1] /summary.lm(result)$coefficients[2,2] )^2
# same as
#  (summary.lm(result)$coefficients[2,3])^2
```


```{r}
model.h0 <- lm(Y ~ 1)
# MC estimation of the permutational distribution of the test statistics
T.beta.sq <- numeric(B)
T.beta.std <- numeric(B)

set.seed(seed)
for(perm in 1:B){
  permutation <- sample(n)
  Y.perm.glob <- model.h0$fitted.values + model.h0$residuals[permutation] 
  # alternatively
  # Y.perm.glob <- Y[permutation]
  perm.fit <- lm(Y.perm.glob ~ x1)
  T.beta.sq[perm] <- (summary.lm(perm.fit)$coefficients[2,3])^2
}

sum(T.beta.sq >= T0.beta.sq)/B
sum(T.beta.std >= T0.beta.std)/B
```


### Multiple regression


Now, we have more than one continuous regressor.

```{r}
# covariate values
x1 <- runif(n,0,10)
x2 <- (1:n)/5
x3 <- rnorm(n,5,5)


# generating model
b0 <- 2
b1 <- 3
b2 <- -2
b3 <- 0
Y <- b0 + b1*x1 + b2*x2 + b3*x3 + stabledist::rstable(n,1.2,0)
```

Let's run some plots...

```{r}
plot(x1,Y,pch=16)
plot(x2,Y,pch=16)
plot(x3,Y,pch=16)
```


And of course due to the strong collinearity we see visually, we expect the test will be significat for at least one regressor.


And, let's see how parametric inference behaves in this case (spoiler alert, badly. Homework: why?)^[If you clicked this footnote, perhaps you were looking for the solution. What was the main message of Lab$03$?]

```{r}
# parametric inference
result <- lm(Y ~ x1 + x2 + x3)
summary(result)
```

We notice that the hypothesis of the model do not hold, in fact we
reject the normality of the residuals:

```{r}
shapiro.test(result$residuals)$p
qqnorm(result$residuals)
```

How do I behave in this case, permutationally?

Let's start with a **global test**.
We have:
$$
H_0: \beta_1 = \beta_2 = \beta_3 = 0 ; \; \implies y = \beta_0 + \epsilon
$$
vs. 
$$
\; H_1: \exists \, l \in \{1, ..., L\}, \; s.t. \beta_l \neq0
$$
(in this exercise $L=3$).
My test statistic is the $F$ statistic:
$$
T = \frac{SS_{reg}}{SS_{res}}
$$
You need not learn this formula by heart. What is important for you to know is that this statistic increases as the Sum of Squares of the reqression w.r.t the sum of squares of the residuals.

An alternative could be the adjusted R-squared statistic (as we saw before you can extract from the summary it after fitting an $lm$ model), which measures the percentage of variability of the dependent variable explained by the model with an adjustment for degrees of freedom^[Again you just have to know it could be a good choice for this scenario. For further details I refer you again to the book of Applied statistics.]

```{r}
T0_glob <- summary(result)$f[1]
T0_glob
```

The permutation scheme to use for the global model is basically, to **permute the responses**...
Basically, if there was no model (i.e. my $H_0$, that every coefficient is 0), it wouldn't matter which input I'm giving, I should expect the same response (see the last lab for the computational proof). So permuting them wouldn't lead to a difference under the null hypothesis. Indeed, under $H_0$:


```{r}
T_H0glob <- numeric(B)

for(perm in 1:B){
  permutation <- sample(n)
  
  Y.perm.glob <- Y[permutation]
  T_H0glob[perm] <- summary(lm(Y.perm.glob ~ x1 + x2 + x3))$f[1]
}

sum(T_H0glob>=T0_glob)/B
```

Ok, the model is significant, let's go ahead with the other tests...
The three test statistics are the absolute value of their t_statistic normalised by their standard error:
$$
T_l = \bigg |\frac{\hat{\beta}_l}{SE_{\hat{\beta}_l}}\bigg|
$$

```{r}
T0_x1 <- abs(summary(result)$coefficients[2,1]) # use coefficients[2,1] for the value
T0_x1

T0_x2 <- abs(summary(result)$coefficients[3,3])
T0_x2

T0_x3 <- abs(summary(result)$coefficients[4,3])
T0_x3
```

And, let's compute the residuals under $H_0$ for the three hypotheses

```{r}

regr.H01 <- lm(Y ~ x2 + x3)
residuals.H01 <- regr.H01$residuals

regr.H02 <- lm(Y ~ x1 + x3)
residuals.H02 <- regr.H02$residuals

regr.H03 <- lm(Y ~ x1 + x2)
residuals.H03 <- regr.H03$residuals
```

Now, let's compute the distribution

```{r}
 T_H01 <- T_H02 <- T_H03 <- numeric(B)

for(perm in 1:B){
  permutation <- sample(n)
  
  residuals.H01.perm <- residuals.H01[permutation]
  Y.perm.H01 <- regr.H01$fitted + residuals.H01.perm
  T_H01[perm] <- abs(summary(lm(Y.perm.H01 ~ x1 + x2 + x3))$coefficients[2,3])
  
  residuals.H02.perm <- residuals.H02[permutation]
  Y.perm.H02 <- regr.H02$fitted + residuals.H02.perm
  T_H02[perm] <- abs(summary(lm(Y.perm.H02 ~ x1 + x2 + x3))$coefficients[3,3])
  
  residuals.H03.perm <- residuals.H03[permutation]
  Y.perm.H03 <- regr.H03$fitted + residuals.H03.perm
  T_H03[perm] <- abs(summary(lm(Y.perm.H03 ~ x1 + x2 + x3))$coefficients[4,3])
  
}

sum(T_H01>=T0_x1)/B
sum(T_H02>=T0_x2)/B
sum(T_H03>=T0_x3)/B
```
I know it seems rather complex because we have to fit the model under $H_1$ to obtain the value of the test statistic, and the model under $H_0$ to estimate residuals and fitted values. So I decided to write cleanly the algorithm to make it clearer.


#### Parenthesis: a possible use of the DDplot

In Lab $01$ we saw the Depth-Depth plot, which is a Nonparametric explorative visual tool to compare two populations. It is actually a nonparametric generalisation of the qqplot, however:
* It also works with multivariate data
* It uses depths that are computed w.r.t the available sample. So we do not compare the empirical depths (which replace the quantiles) with theoretical ones, but with the depths of another sample.

The interpretation is the same as in the qqplot. If we see approximately a straight line, it means that both (empirical) distributions are similar.

Thus, if we wanted to see the e.c.d.f. of the residuals w.r.t a normal one, we would have to build a DD-plot where one sample is the one of the residuals, and the other one of a normal distribution.

(Note that to make them comparable, I sampled from a normal distribution with the same variance as the residuals, which is in turn estimated in the $lm$ object.)

```{r}
n <- length(result$residuals)
set.seed(20)
normal <- as.matrix(rnorm(n, sd=summary.lm(result)$sigma))

DepthProc::ddPlot(x = as.matrix(normal), # use as.matrix since ddplot usually receives multivariate data
                  y = as.matrix(as.matrix(result$residuals)),depth_params = list(method='Tukey'))
```
We notice that the points go far away from the $y=x$ line.
Naturally, if we compared two samples from the same normal distribution, we would have a very high correlation in the depths:
```{r}
normal <- as.matrix(rnorm(n, 
                          sd=summary.lm(result)$sigma)) 
normal2 <- as.matrix(rnorm(n, 
                          sd=summary.lm(result)$sigma))


DepthProc::ddPlot(x = normal, y = normal2 ,depth_params = list(method='Tukey'))
```
And the conclusion we gather from such DD-plot is that the samples are indeed quite likely to be from the same distribution.


## Independence test

There is another permutation test we can perform.

We have a sample of bivariate data $(\mathbf{x}, \mathbf{y}) \sim (X,Y)$.

For simplicity, let us sample from a bivariate normal distribution

```{r}
SIGMA <-  diag(2)*2
data <- MASS::mvrnorm(30, c(0,0), SIGMA)
plot(data)
```

We test: 
$$
H_0: X \!\perp\!\!\!\perp Y \,\,\, versus \,\,\, H_1: X \not\!\perp\!\!\!\perp Y
$$
and the test statistic is the samples squared correlation coefficient.

Under $H_0$, they are independent, so **within each vector**, we can permute the statistical units.

```{r}

T0 <- cor(data[,1], data[,2])^2
n <- nrow(data)

T0.perm.distrib <- numeric(B)

for(perm in 1:B){
  permutation.x <- sample(n)
  permutaiton.y <- sample(n)
  T0.perm.distrib[perm] <- cor(data[permutation.x,1], 
                                 data[permutaiton.y,2])^2
  }

sum(T0.perm.distrib >=T0)/B

```






